Graph Topological Aspects of Granger Causal Network Learning\\ \large Supplementary Material

### 2019-07-18 paper_theory/proofs.tex
Better proofs.

\documentclass{statsoc}
\usepackage{fullpage}
\usepackage{ccfonts,eulervm}
\usepackage{tabularx}
\usepackage{xr}
\usepackage{chngcntr}
\usepackage{apptools}
\AtAppendix{\counterwithin{lemma}{section}}
\AtAppendix{\counterwithin{definition}{section}}
\externaldocument{paper_theory}

@@

@@

\section{Proofs}
\subsection{Ancillary Results}
\label{apx:ancillary_results}
\begin{theorem}[Granger Causality Equivalences \ref{thm:granger_causality_equivalences}]
  The following are equivalent:

@@

  $(a) \Rightarrow (b)$ follows as a result of the uniqueness of orthogonal
  projection (i.e. the best estimate is necessarily the coefficients
  of the model).  $(b) \Rightarrow (c)$ follows since in computing
  $(y - \linE{y}{\H_{t - 1}^{-j}})$ for $y \in H_t^i$ it is sufficient
  to consider $y = x_i(t)$ by linearity, then since
  $H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$ we have
  $(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$ since
  $\B_{ij}(z) = 0$ and $v_i(t) \perp \H_{t - 1}$.  $(c) \iff (d)$ is a result
  of the equivalence in Definition
  \ref{lem:conditional_orthogonality_equivalence}.  And,
  $(d) \implies (a)$ follows directly from the Definition.
\end{proof}

@@

This statement is ultimately about the sparsity pattern in the Wold
decomposition matrices $A(\tau)$ since
$x_i(t) = \sum_{\tau = 0}^\infty \sum_{j = 1}^n A_{ij}(\tau)v_j(t -
\tau)$.  The proposition states that if $j \not \in \anc{i}$ then
$\A_{ij}(z) = 0$.

\begin{proof}
  Equation \eqref{eqn:parent_expansion} is immediate from the
  $\VAR(\infty)$ representation of \eqref{eqn:ar_representation} and
  Theorem \ref{thm:granger_causality_equivalences}, we are left to
  demonstrate \eqref{eqn:ancestor_expansion}.

@@

  where $(I - \B(z))^{-1} = \A(z)$ due to the uniqueness of
  (\ref{eqn:wold}).  Since $\B(z)$ is stable we have

@@

  \begin{align*}
    x_i(t) &= [(I - \B(z))^{-1}v(t)]_i = \sum_{j = 1}^n \A_{ij}(z) v_j(t)\\
           &= \A_{ii}(z) v_i(t) + \sum_{\substack{j \in \anc{i} \\ j \ne i}} \A_{ij}(z) v_j(t)
  \end{align*}
\end{proof}

@@

  then $\H_t^{(i)} \perp \H_t^{(j)}$, that is,
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  Moreover,
  this means that $j \npwgc i$ and $\linE{x_j(t)}{\H_{t - 1}^i} = 0$.
\end{proposition}

\begin{proof}
  We show directly that
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  To this end,
  fix $s, \tau \ge 0$, then by expanding with equation \eqref{eqn:ancestor_expansion} we have

  \begin{align*}
    \E x_i(t - s)x_j(t - \tau)
    &= \E \big(\A_{ii}(z)v_i(t - s)\big)\big(\A_{jj}(z)v_j(t - \tau)\big)\\
    &+ \sum_{\substack{k \in \anc{i} \\ k \ne i}}\E[\big(\A_{ik}(z)v_k(t - s)\big)\big(\A_{jj}(z)v_j(t - \tau)\big)]\\
    &+ \sum_{\substack{k \in \anc{j} \\ k \ne j}}\E[\big(\A_{ii}(z)v_i(t - s)\big) \big(\A_{jk}(z) v_k(t - \tau)\big)]\\
    &+ \sum_{\substack{k \in \anc{i} \\ k \ne i}}\sum_{\substack{\ell \in \anc{j} \\ \ell \ne j}}\E[\big(\A_{ik}(z)v_k(t - s)\big)\big(\A_{j\ell}(z)v_\ell(t - \tau)\big)].
  \end{align*}

@@

@@

  \begin{equation*}
    \inner{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^i}}{x_j(t) - \linE{x_j(t)}{\H_{t - 1}^i}} = 0.
  \end{equation*}

  which by the orthogonality principle is equivalent to

  \begin{equation}
    \label{eqn:proof_inner0}
    \inner{x_i(t)}{x_j(t) - \linE{x_j(t)}{\H_{t - 1}^i}} = 0.
  \end{equation}

  Define the disjoint sets

  \begin{align*}
    C_0(u) &= \{k \in \pa{u}\ |\ i \not\in \anc{k}, k \ne i \}\\
    C_1(u) &= \{k \in \pa{u}\ |\ i \in \anc{k} \text{ or } k = i \}.
  \end{align*}

  We can then expand the parents of $j$ using Equation \ref{eqn:parent_expansion} as

  \begin{equation*}
    x_j(t) = v_j(t) + \sum_{k \in C_0(j)} B_{jk}(z) x_k(t) + \sum_{k \in C_1(j)} B_{jk}(z) x_k(t),
  \end{equation*}

  which when substituted into the left hand side of Equation
  \ref{eqn:proof_inner0} results (by Proposition
  \ref{prop:separated_ancestor_uncorrelated}) in

  \begin{equation*}
    \inner{x_i(t)}{\sum_{k \in C_1(j)} B_{jk}(z) x_k(t) - \sum_{k \in C_1(j)} \linE{B_{jk}(z) x_k(t)}{\H_{t - 1}^i}}.
  \end{equation*}

  We can continue this process recursively (i.e. split each
  $k \in C_1(j)$ into $C_0(k)$ and $C_1(k)$) which must eventually
  terminate with $C_1(u) = \{i\}$.  Therefore, there exists some
  causal filter $\Phi(z)$ such that Equation \ref{eqn:proof_inner0} is
  equivalent to

  \begin{equation}
    \label{eqn:proof_inner0}
    \inner{x_i(t)}{\Phi(z)x_i(t) - \linE{\Phi(z)x_i(t)}{\H_{t - 1}^i}},
  \end{equation}

  which is $0$ since $\linE{\Phi(z)x_i(t)}{\H_{t - 1}^i} = \Phi(z)x_i(t)$.
\end{proof}

@@

  Firstly, notice that every $u \in \big(\pa{j} \setminus \{i\}\big)$
  necessarily inherits these same two properties.  This follows since
  if we also had $u \in \anc{i}$ then $u \in \anc{i} \cap \anc{j}$ so that every
  $u \rightarrow \cdots \rightarrow j$ path must contain $i$, but
  $u \in \pa{j}$, so this is not the case since $u \rightarrow j$ is a
  path that doesn't contain $i$; moreover, if we consider
  $w \in \anc{i} \cap \anc{u}$ then we also have
  $w \in \anc{i} \cap \anc{j}$ so every
  $w \rightarrow \cdots \rightarrow j$ path must contain $i$.  These
  properties therefore extend inductively to every
  $u \in \big(\anc{j} \setminus \{i\}\big)$.

@@

  We notice that for any $u$ having the properties $(a), (b)$ above, we
  must have $C_2(u) = \emptyset$ since if $k \in C_2(u)$ then
  $\exists w \in \anc{i} \cap \anc{k}$ s.t. $i \not \in \anc{k}$ and
  therefore there must be a path $\gcgpath{w}{k} \rightarrow u$ which
  does not contain $i$.

  Using this partition, we will expand $x_j(t)$ in terms of it's
  parents, and recursively expand nodes in $C_1$ until we reach a case
  where $C_1 = \emptyset$.  For the first step equation
  \eqref{eqn:parent_expansion} gives us:

  \begin{equation}
    \label{eqn:xj_partition_expansion}
    x_j(t) = \A_{jj}(z)\Big(v_j(t) + \sum_{k \in C_0(j)}\B_{ik}(z)x_k(t) + \sum_{k \in C_1(j)}\B_{ik}(z)x_k(t)\Big).
  \end{equation}

  Using this representation we choose an arbitrary $\Phi(z) x_i(t - 1) \in \H_{t - 1}^{(j)}$ and show that

  \begin{equation}
    \inner{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}}{\Phi(z)x_j(t - 1) - \linE{\Phi(z)x_j(t - 1)}{\H_{t - 1}^{(i)}}} = 0,
  \end{equation}

  which will imply (by Theorem \ref{thm:granger_causality_equivalences}) that $j \npwgc i$ and for which it is equivalent to show that

  \begin{equation}
    \label{eqn:sufficient_inner_prod}
    \inner{x_i(t)}{\Phi(z)x_j(t - 1) - \linE{\Phi(z)x_j(t - 1)}{\H_{t - 1}^{(i)}}} = 0,
  \end{equation}

  by the orthogonality principle since $\linE{x_i(t)}{\H_{t - 1}^{(i)}} \in \H_{t - 1}^{(i)}$.  Substituting (\ref{eqn:xj_partition_expansion}) into (\ref{eqn:sufficient_inner_prod}) and starting with the first term we have

  \begin{align*}
    &\inner{x_i(t)}{\Phi(z)\A_{jj}(z)v_j(t - 1) - \linE{\Phi(z)\A_{jj}(z)v_j(t - 1)}{\H_{t - 1}^{(i)}}}\\
    \overset{(\alpha)}{=}\ &\inner{\A_{ii}(z)v_i(t) + \sum_{k \in \anc{i}}\A_{ik}(z)v_k(t)}{\Phi(z)\A_{jj}(z)v_j(t - 1)}\\
    \overset{(\beta)}{=}\ &0,
  \end{align*}

  where $(\alpha)$ follows by expanding $x_i(t)$ with (\ref{eqn:ancestor_expansion}) and $\linE{\Phi(z)\A_{jj}(z)v_j(t - 1)}{\H_{t - 1}^{(i)}} = 0$ because $\forall \tau, s$

  \begin{equation*}
    \E v_j(t - \tau) x_i(t - s) = \E v_j(t - \tau) \sum_{k \in \anc{i} \cup \{i\}}\A_{ik}(z)v_k(t - s) = 0,
  \end{equation*}

  since $j \not \in \anc{i}$; $(\beta)$ follows similarly, that is, $j \not \in \anc{i}$.  Secondly we see that $\forall k \in C_0(j)$

  \begin{equation*}
    \inner{x_i(t)}{\Phi(z)\A_{jj}(z)\B_{ik}(z)x_k(t - 1) - \linE{\Phi(z)\A_{jj}(z)\B_{ik}(z)x_k(t - 1)}{\H_{t - 1}^{(i)}}} = 0,
  \end{equation*}

  which follows from Proposition \ref{prop:ancestor_uncorrelated}.  Finally,
  for $k \in C_1(j)$ the case $k = i$ is immediate (since the error in
  estimating $\Phi(z)\B_{ii}(z)x_i(t)$ given $\H_{t - 1}^{(i)}$ is
  $0$), so suppose $k \ne i$.  We know from above that $k$ inherits the
  key properties referred to as $(a)$ and $(b)$ above and therefore we
  can recursively expand $k$ in the same way as in equation
  (\ref{eqn:xj_partition_expansion}).  Continuing this recursion for
  each $k \in C_1(j)$ (where $k \ne i$) must eventually terminate since
  $i \in \anc{k}$.
\end{proof}

\begin{proposition}
  \label{prop:sc_graph_common_anc}
  In a strongly causal graph if $j \in \anc{i}$ then any
  $k \in \anc{i} \cap \anc{j}$ is not a confounder, that is,
  the unique path from $k$ to $i$ contains $j$.
\end{proposition}
\begin{proof}
  Suppose that there is a path from $k$ to $i$ which does not contain
  $j$.  In this case, there are multiple paths from $k$ to $i$ (one of
  which \textit{does} go through $j$ since $j \in \anc{i}$) which
  contradicts the assumption of strong causality.
\end{proof}

@@

  % Should this be written as a lemma?
  \begin{align*}
    x_i(t) &= v_i(t) + \B_{ii}(z) x_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z) x_k(t)\\
           &= \underbrace{v_{a_0}(t) + \B_{a_0a_0}(z) x_i(t) + \sum_{\substack{k \in \pa{a_0} \\ k \ne a_1}}\B_{a_0 k}(z) x_k(t)}_{\defeq \wtalpha{a_0}{a_1}} + \B_{a_0a_1}(z)x_{a_1}(t)\\
           &= \wtalpha{a_0}{a_1} + \B_{a_0a_1}(z)\big[\wtalpha{a_1}{a_2} + \B_{a_1a_2}(z)x_{a_2}(t) \big]\\
           &\overset{(a)}{=} \sum_{\ell = 0}^r \underbrace{\Big(\prod_{m = 0}^{\ell - 1} \B_{a_m a_{m + 1}}(z) \Big)}_{\defeq F_\ell(z)} \wtalpha{a_\ell}{a_{\ell + 1}} + \Big(\prod_{m = 0}^{r}\B_{a_m a_{m + 1}}(z)\Big)x_{a_{r + 1}}(t)\\
           &= \sum_{\ell = 0}^r F_\ell(z) \wtalpha{a_\ell}{a_{\ell + 1}} + F_{r + 1}(z) x_j(t)
  \end{align*}

@@

  or by rearranging and applying the representation obtained earlier, if and only if

  \begin{equation*}
    x_i(t) \overset{\text{a.s.}}{=} \linE{F_{r + 1}(z)x_j(t)}{\H_{t - 1}^{(i)}},
  \end{equation*}

  but this is impossible since $x_i(t) \not \in \H_{t - 1}^{(i)}$.
\end{proof}

\begin{lemma}
  \label{lem:time_lag_cancellation}
  Suppose $v(t)$ is a scalar sequence with unit variance and zero
  autocorrelation and let $\A(z), \B(z)$ be nonzero and strictly
  causal (i.e. $\tau_0(\A) \ge 1$) linear filters.  Then,

@@

  due to the uncorrelatedness assumptions on $v(t)$.  This expression
  is $0$ if and only if $\tau_0(\A) \ge 1 + \tau_\infty(\B)$ or if
  $\tau_0(\B) \ge 1 + \tau_\infty(\A)$ or if the coefficients are
  orthogonal along the common support.

  Specializing this fact to $\inner{F(z)\A(z)v(t)}{\B(z)v(t)} = 0$ we
  see that the coefficients cannot be orthogonal for every choice of
  $F$, and that $\text{sup}_F \tau_\infty(F\A) = \infty$, leaving only
  the possibility that

  \begin{align*}
    \tau_0(F\A) \ge 1 + \tau_\infty(\B) \forall F &\overset{(a)}{\iff} \tau_0(\A) \ge 1 + \tau_\infty(\B) - \underset{F}{\text{min }} \tau_0(F)\\
    &\overset{(b)}{\iff} \tau_0(\A) \ge \tau_\infty(\B),
  \end{align*}

@@

\begin{proposition}
  \label{prop:persistence_converse}
  Suppose $\gcg$ is a strongly causal DAG and that $x(t)$ is
  persistent, then if there exists a $k$ which confounds $(i, j)$ we
  have $i \pwgc j$ and $j \pwgc i$.
\end{proposition}
\begin{proof}
  We will show that $j \pwgc i$, the other being symmetric.  First
  note also that by proposition \ref{prop:sc_graph_common_anc} we
  cannot have $i \in \anc{j}$ or $j \in \anc{i}$ and therefore every
  $k \in \anc{i}\cap\anc{j}$ will be a confounder.

>>>>

\documentclass{statsoc}
\usepackage{fullpage}
\usepackage{ccfonts,eulervm}
\usepackage{tabularx}
\usepackage{xr}
\usepackage{chngcntr}
\usepackage{apptools}
\AtAppendix{\counterwithin{lemma}{section}}
\AtAppendix{\counterwithin{definition}{section}}
\AtAppendix{\counterwithin{equation}{section}}
\externaldocument{paper_theory}

@@

References to equations, lemmas, etc., in this document are prefixed
with their section, whereas prefix-free equation numbers refer to the
main document.  e.g. ``Equation (1)'' refers to the first equation in
the main document, and ``Equation (A.1)'' refers to the first equation
in this document.

@@

\section{Proofs}
\subsection{Preparatory Results}
\label{apx:ancillary_results}
\begin{theorem}[Granger Causality Equivalences \ref{thm:granger_causality_equivalences}]
  The following are equivalent:

@@

  $(a) \Rightarrow (b)$ follows as a result of the uniqueness of orthogonal
  projection (i.e. the best estimate is necessarily the coefficients
  of the model).  $(b) \Rightarrow (c)$ follows since in computing
  $(y - \linE{y}{\H_{t - 1}^{-j}})$ for $y \in H_t^i$ it is sufficient
  to consider $y = x_i(t)$ by linearity, then since
  $H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$ we have
  $(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$ since
  $\B_{ij}(z) = 0$.  The result follows since
  $v_i(t) \perp \H_{t - 1}$.  $(c) \iff (d)$ is a result of the equivalence
  in Definition \ref{lem:conditional_orthogonality_equivalence}.  And,
  $(d) \implies (a)$ follows directly from the Definition.
\end{proof}

@@

\begin{proof}
  Equation \eqref{eqn:parent_expansion} is immediate from the
  $\VAR(\infty)$ representation of Equation \eqref{eqn:ar_representation}
  and Theorem \ref{thm:granger_causality_equivalences}, we are left to
  demonstrate \eqref{eqn:ancestor_expansion}.

@@

  where necessarily $(I - \B(z))^{-1} = \A(z)$ due to the uniqueness
  of the Wold decomposition.  Since $\B(z)$ is stable we have

@@

  \begin{align*}
    x_i(t) &= [(I - \B(z))^{-1}v(t)]_i\\
           &= \sum_{j = 1}^n \A_{ij}(z) v_j(t)\\
           &= \A_{ii}(z) v_i(t) + \sum_{\substack{j \in \anc{i} \\ j \ne i}} \A_{ij}(z) v_j(t)
  \end{align*}
\end{proof}

@@

  then $\H_t^{(i)} \perp \H_t^{(j)}$, that is,
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  Moreover,
  this means that $j \npwgc i$ and $\linE{x_j(t)}{\H_t^i} = 0$.
\end{proposition}

\begin{proof}
  We show directly that
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  To this end, fix
  $s, \tau \ge 0$, then by expanding with Equation
  \eqref{eqn:ancestor_expansion} we have

  \begin{align*}
    \E x_i(t - s)x_j(t - \tau)
    &= \E \big(\A_{ii}(z)v_i(t - s)\big)\big(\A_{jj}(z)v_j(t - \tau)\big)\\
    &+ \sum_{\substack{k \in \anc{i} \\ k \ne i}}\E[\big(\A_{ik}(z)v_k(t - s)\big)\big(\A_{jj}(z)v_j(t - \tau)\big)]\\
    &+ \sum_{\substack{\ell \in \anc{j} \\ \ell \ne j}}\E[\big(\A_{ii}(z)v_i(t - s)\big) \big(\A_{j\ell}(z) v_\ell(t - \tau)\big)]\\
    &+ \sum_{\substack{k \in \anc{i} \\ k \ne i}}\sum_{\substack{\ell \in \anc{j} \\ \ell \ne j}}\E[\big(\A_{ik}(z)v_k(t - s)\big)\big(\A_{j\ell}(z)v_\ell(t - \tau)\big)].
  \end{align*}

@@

\begin{lemma}
  \label{lem:vj_perp}
  Consider distinct nodes $i, j$ in a Granger-causality graph $\gcg$.
  If $j \not \in \anc{i}$, then $\H_t^{v_j} \perp \H_{t}^{i}$, and therefore
  for any causal filter $\Phi(z)$ we have

  \begin{align*}
    \linE{\Phi(z)v_j(t)}{\H_{t - 1}^i} &= 0\\
    \inner{x_i(t)}{\Phi(z)v_j(t)} & = 0.
  \end{align*}
\end{lemma}
\begin{proof}
  Fix $\tau, s \ge 0$, then by expanding with Equation \ref{eqn:ancestor_expansion}

  \begin{align*}
    \E[x_i(t - \tau)v_j(t - s)] &= \E[\big(\A_{ii}(z)v_i(t) + \sum_{k \in \anc{i}}\A_{ik}(z)v_k(t) \big) v_j(t - s)]\\
    &= 0.
  \end{align*}

  This follows since $i \ne j$ and $j \not \in \anc{i}$ and $v(t)$ is
  isotrophic and uncorrelated.
\end{proof}

@@

  \begin{equation*}
    \forall \psi \in \H_{t - 1}^j\ \inner{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^i}}{\psi - \linE{\psi}{\H_{t - 1}^i}} = 0.
  \end{equation*}

  which by the orthogonality principle and by representing
  $\psi \in \H_{t - 1}^j$ via the action of some strictly causal filter
  $\Phi(z)$ on $x_j(t)$ is equivalent to

  \begin{equation}
    \label{eqn:proof_inner0}
    \inner{x_i(t)}{\Phi(z)x_j(t) - \linE{\Phi(z)x_j(t)}{\H_{t - 1}^i}} = 0.
  \end{equation}

  If we expand $x_j(t)$ using Equation \eqref{eqn:ancestor_expansion},
  the left hand side of \eqref{eqn:proof_inner0} becomes

  \begin{equation*}
    \inner{x_i(t)}{\sum_{k \in \anc{j} \cup \{j\}} \Big(\Phi(z)\A_{jk}(z)v_k(t) - \linE{\Phi(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^i}\Big)}.
  \end{equation*}

  We see that this is $0$ by Lemma \ref{lem:vj_perp} since
  $j \not \in \anc{i}$, and
  \[
    \anc{i} \cap \anc{j} = \emptyset \implies \forall k \in \anc{j}: k \not \in \anc{i}.
  \]
\end{proof}

% I believe this is an incorrect proof since it relies on inverting
% (1 - B_{jj}(z)), which need not be invertible.  For example,
% B(z) = [[r, -a], [a, 0]]z^{-1} is invertible whenever
% max(|r - root(r^2 - 4a^2)|, |r + root(r^2 - 4a^2)|) < 2
% which includes values of r > 1.

% \begin{proof}
%   By Theorem \ref{thm:granger_causality_equivalences} it suffices to show that

%   \begin{equation*}
%     \forall \psi \in \H_{t - 1}^j\ \inner{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^i}}{\psi - \linE{\psi}{\H_{t - 1}^i}} = 0.
%   \end{equation*}

%   which by the orthogonality principle and by representing
%   $\psi \in \H_{t - 1}^j$ via the action of some strictly causal filter
%   $\Phi(z)$ on $x_j(t)$ is equivalent to

%   \begin{equation}
%     \label{eqn:proof_inner0}
%     \inner{x_i(t)}{\Phi(z)x_j(t) - \linE{\Phi(z)x_j(t)}{\H_{t - 1}^i}} = 0.
%   \end{equation}

%   Given a node $u$, consider the partition of $\pa{u}$:

%   \begin{align*}
%     C_0(u) &= \{k \in \pa{u}\ |\ i \not\in \anc{k}, k \ne i \}\\
%     C_1(u) &= \{k \in \pa{u}\ |\ i \in \anc{k} \text{ or } k = i \}.
%   \end{align*}

%   We can then expand the parents of $j$ using Equation \ref{eqn:parent_expansion} as

%   \begin{align*}
%     x_j(t) &= v_j(t) + \B_{jj}(z)x_j(t) + \sum_{k \in C_0(j)} B_{jk}(z) x_k(t) + \sum_{k \in C_1(j)} B_{jk}(z) x_k(t)\\
%     &= \widetilde{A}_{jj}(z)\Big(v_j(t) + \sum_{k \in C_0(j)} B_{jk}(z) x_k(t) + \sum_{k \in C_1(j)} B_{jk}(z) x_k(t)\Big),
%   \end{align*}

%   % \todo{Is this true??}
%   where $(1 - B_{jj}(z))^{-1} = \widetilde{A}_{jj}(z)$, which exists
%   due to the invertibility of $\A(z)$. When this expression is
%   substituted into the left hand side of Equation
%   \ref{eqn:proof_inner0} it results in

%   \begin{equation}
%     \label{eqn:proof_inner1}
%     \begin{aligned}
%       \inner{x_i(t)}{\sum_{k \in C_1(j)} \big(\widetilde{A}_{jj}(z)\Phi(z)B_{jk}(z) x_k(t) &- \linE{\widetilde{A}_{jj}(z)\Phi(z)B_{jk}(z) x_k(t)}{\H_{t - 1}^i}\big)},
%     \end{aligned}
%   \end{equation}

%   where (by Proposition \ref{prop:separated_ancestor_uncorrelated}) we
%   have eliminated each of the terms from $C_0(u)$, and by Lemma
%   \ref{lem:vj_perp} the terms involving $v_j(t)$.

%   Now, for any $u \in \anc{j}$, we must have $u \not \in \anc{i}$, since
%   otherwise $u \in \anc{i} \cap \anc{j}$, which is empty by hypothesis.
%   Moreover, for any $u \in \anc{j}$, we have
%   $\anc{i} \cap \anc{u} = \emptyset$, since
%   $w \in \anc{i} \cap \anc{u} \implies w \in \anc{i} \cap \anc{j}$. Therefore,
%   we can continue the above process recursively splitting each
%   $k \in C_1(j)$ into $C_0(k)$ and $C_1(k)$) upwards along each path
%   from $j$ to $i$, terminating whenever we encounter $i$.  This process
%   is guaranteed to terminate since $i \in \anc{k}$.  This leaves us with
%   some strictly causal filter $\Psi(z)$ such that Expression
%   \ref{eqn:proof_inner1} is equivalent to

%   \begin{equation}
%     \inner{x_i(t)}{\Psi(z)x_i(t) - \linE{\Psi(z)x_i(t)}{\H_{t - 1}^i}},
%   \end{equation}

%   which is $0$ since $\Psi(z)x_i(t) \in \H_{t - 1}^i$ and
%   therefore $\linE{\Psi(z)x_i(t)}{\H_{t - 1}^i} = \Psi(z)x_i(t)$.
% \end{proof}

@@

  Firstly, notice that every $u \in \big(\pa{j} \setminus \{i\}\big)$
  necessarily inherits these same two properties.  This follows since
  if we also had $u \in \anc{i}$ then $u \in \anc{i} \cap \anc{j}$ so by our
  assumption every $u \rightarrow \cdots \rightarrow j$ path must contain
  $i$, but $u \in \pa{j}$ so $u \rightarrow j$ is a path that doesn't contain
  $i$, and therefore $u \not\in \anc{i}$; moreover, if we consider
  $w \in \anc{i} \cap \anc{u}$ then we also have
  $w \in \anc{i} \cap \anc{j}$ so the assumption implies that every
  $w \rightarrow \cdots \rightarrow j$ path must contain $i$.  These properties therefore
  extend inductively to every $u \in \big(\anc{j} \setminus \{i\}\big)$.

@@

  We notice that for any $u$ having the properties $(a), (b)$ above,
  we must have $C_2(u) = \emptyset$ since if $k \in C_2(u)$ then
  $\exists w \in \anc{i} \cap \anc{k}$ (and
  $w \in \anc{i} \cap \anc{u}$, since $k \in \pa{u}$) such that
  $i \not \in \anc{k}$ and therefore there must be a path
  $\gcgpath{w}{k} \rightarrow u$ which does not contain $i$, contradicting
  property $(b)$.

  Using this partition, and in a similar fashion to the proof of
  Proposition \ref{prop:ancestor_uncorrelated}, we will expand
  $x_j(t)$ in terms of it's parents, and recursively expand nodes in
  $C_1$ until reaching $i$.  For the first step equation
  \eqref{eqn:parent_expansion} gives us:

  \begin{equation}
    \label{eqn:xj_partition_expansion}
    x_j(t) = \widetilde{\A}_{jj}(z) \Big(v_j(t) + \sum_{k \in C_0(j)} \B_{ik}(z)x_k(t) + \sum_{k \in C_1(j)}\B_{ik}(z)x_k(t)\Big).
  \end{equation}

  Using this representation we choose an arbitrary element of
  $\H_{t - 1}^j$ and represent it as the action of a strictly causal
  filter $\Phi(z)$, i.e.  $\Phi(z) x_j(t) \in \H_{t - 1}^{(j)}$ and show that

  \begin{equation}
    \label{eqn:sufficient_inner_prod}
    \inner{x_i(t)}{\widetilde{\A}_{jj}(z)\Phi(z)x_j(t) - \linE{\widetilde{\A}_{jj}(z)\Phi(z)x_j(t)}{\H_{t - 1}^{(i)}}} = 0,
  \end{equation}

  which will imply (by Theorem
  \ref{thm:granger_causality_equivalences}) that $j \npwgc i$.

  Substituting (\ref{eqn:xj_partition_expansion}) into
  (\ref{eqn:sufficient_inner_prod}) we have cancellation of all terms
  in which $k \in C_0(j)$ by Proposition \ref{prop:ancestor_properties},
  and cancellation of the term involving $v_j(t)$ by Lemma \ref{lem:vj_perp}.

  We are therefore left with

  \begin{equation}
    \label{eqn:refwornv}  % An arbitrary code for reference in just one place
    \inner{x_i(t)}{\sum_{k \in C_1(j)} \big(\widetilde{\A}_{jj}(z)\Phi(z) \B_{ik}(z)x_k(t) - \linE{\widetilde{\A}_{jj}(z)\Phi(z) \B_{ik}(z)x_k(t)}{\H_{t - 1}^i}\big)}.
  \end{equation}

  Since each such $k$ inherits the needed properties $(a), (b)$ we can
  again (as in Proposition \ref{prop:ancestor_properties}) recursively
  expand this expression until we reach $i$, obtaining a strictly
  causal $\Psi(z)$ such that Expression \eqref{eqn:refwornv} is equal to

  \begin{equation}
    \inner{x_i(t)}{\Psi(z)x_i(t) - \linE{\Psi(z)x_i(t)}{\H_{t - 1}^i}} = 0.
  \end{equation}
\end{proof}

\begin{proposition}
  \label{prop:sc_graph_common_anc}
  In a strongly causal graph if $j \in \anc{i}$ then any
  $k \in \anc{i} \cap \anc{j}$ is not a confounder, that is,
  the unique path from $k$ to $i$ contains $j$.
\end{proposition}
\begin{proof}
  Suppose that there is a path from $k$ to $i$ which does not contain
  $j$.  In this case, there are multiple paths from $k$ to $i$ (one of
  which \textit{does} go through $j$, since $j \in \anc{i}$) which
  contradicts the assumption of strong causality.
\end{proof}

@@

  \begin{align*}
    x_i(t) &= v_i(t) + \B_{ii}(z) x_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z) x_k(t)\\
           &= \underbrace{v_{a_0}(t) + \B_{a_0a_0}(z) x_i(t) + \sum_{\substack{k \in \pa{a_0} \\ k \ne a_1}}\B_{a_0 k}(z) x_k(t)}_{\defeq \wtalpha{a_0}{a_1}} + \B_{a_0a_1}(z)x_{a_1}(t)\\
           &= \wtalpha{a_0}{a_1} + \B_{a_0a_1}(z)\big[\wtalpha{a_1}{a_2} + \B_{a_1a_2}(z)x_{a_2}(t) \big]\\
           &\overset{(a)}{=} \sum_{\ell = 0}^r \underbrace{\Big(\prod_{m = 0}^{\ell - 1} \B_{a_m a_{m + 1}}(z) \Big)}_{\defeq F_\ell(z)} \wtalpha{a_\ell}{a_{\ell + 1}} + \Big(\prod_{m = 0}^{r}\B_{a_m a_{m + 1}}(z)\Big)x_{a_{r + 1}}(t)\\
           &= \sum_{\ell = 0}^r F_\ell(z) \wtalpha{a_\ell}{a_{\ell + 1}} + F_{r + 1}(z) x_j(t)
  \end{align*}

@@

  or by rearranging and applying the representation for $x_i(t)$
  obtained earlier, if and only if

  \begin{equation*}
    x_i(t) \overset{\text{a.s.}}{=} \linE{F_{r + 1}(z)x_j(t)}{\H_{t - 1}^{(i)}}.
  \end{equation*}

  But, this is impossible since $x_i(t) \not \in \H_{t - 1}^{(i)}$.
\end{proof}

\begin{lemma}
  \label{lem:time_lag_cancellation}
  Suppose $v(t)$ is a scalar process with unit variance and zero
  autocorrelation and let $\A(z), \B(z)$ be nonzero and strictly
  causal (i.e. $\tau_0(\A) \ge 1$) linear filters.  Then,

@@

  since $\E[v(t - s)v(t - \tau)] = \delta_{s - \tau}$.  This expression is
  $0$ if and only if $\tau_0(\A) \ge 1 + \tau_\infty(\B)$ or if
  $\tau_0(\B) \ge 1 + \tau_\infty(\A)$ or if the coefficients are orthogonal along
  the common support.

  Specializing this fact to $\inner{F(z)\A(z)v(t)}{\B(z)v(t)}$ we
  see that the coefficients cannot be orthogonal for every choice of
  $F$, and that $\text{sup}_F \tau_\infty(F\A) = \infty$, leaving only
  the possibility that

  \begin{align*}
     \forall F\ \tau_0(F\A) \ge 1 + \tau_\infty(\B) &\overset{(a)}{\iff} \tau_0(\A) \ge 1 + \tau_\infty(\B) - \underset{F}{\text{min }} \tau_0(F)\\
    &\overset{(b)}{\iff} \tau_0(\A) \ge \tau_\infty(\B),
  \end{align*}

@@

\begin{proposition}
  \label{prop:persistence_converse}
  Suppose $\gcg$ is a strongly causal DAG and that $x(t)$ is
  persistent, then if there exists a $k$ which confounds $(i, j)$ we
  have $i \pwgc j$ and $j \pwgc i$.
\end{proposition}
\begin{proof}
  We will show that $j \pwgc i$, the other being symmetric.  First
  note that by the contrapositive of proposition
  \ref{prop:sc_graph_common_anc} we cannot have $i \in \anc{j}$ or
  $j \in \anc{i}$ (else $k$ would not be a confounder) and therefore
  \textit{every} $u \in \anc{i}\cap\anc{j}$ will be a confounder.
